Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-28 tagless/tagless-final.tex
Section 4 is mostly done

\section{Our base language: Tagless interpreter}\label{language}

Introduce the language, The language is simply-typed lambda-calculus
with fixpoint, integers, booleans and comparison, whose BNF grammar is

@@

This time we use higher-order abstract syntax (cite: but not
Elliott. Cite Dale Miller!) rather than de Bruijn indices.  The
language is close to the language of Xi03, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Fibonacci, factorial, and power.

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
the term `constructor' functions are defined as methods in a type class
Symantics (the name means that the class interface gives the syntax for
the source language and its instances give the semantics).

@@

We can do the above in (Meta)OCaml as well, replacing type classes with
functors, via a module type.  For this section, the following simple
signature will be sufficient:

@@

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
     if_ (leq n (int 0)) (fun () -> int 1)
         (fun () -> mul x (app self 
                               (add n (int (-1))))))))
 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
The dummy unit argument in |testpowfix| is an artefact of MetaOCaml
(which we will be using to compile our tests in Sections below),
again related to monorphism. To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means, the type
must be polymorhic. The value restriction then dictates 
the definitions of our object terms must look syntactically like
values. Alternatively to giving an |()| argument, we could have used
rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

The first instantiation of the Symantics structure is an interpreter:
evaluating the object (DSL) term to its value in the meta-language. 
The following module essentially interprets (gives one-to-one
correspondence to) each operation of the object language as the
corresponding operation in the meta-language.
\begin{code}
module R  = struct
  type ('c,'dv) repr = 'dv (* absolutely no wrappers *)
  let int  (x:int)  = x
  let bool (b:bool) = b
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then (et ()) else (ee ())

@@

We have also implemented the ``compiler'' in Haskell. Since Haskell
has no (convenient, and typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with the
constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
meaning. Unlike our staging language (which is based on higher-order
abstract syntax), our bytecode, to be realistic, uses integer-named
variables. We then define 
\begin{code}
  newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for the creation of fresh variable
names. We define the compiler by making |C| to be the instance of the
class |Symantics|. The implementation is quite similar (but more
verbose) than the corresponding MetaOCaml code given above. The
accompanying complete code gives more details.

@@

\subsection{Non-functorialness of functions problem}
the sv type problem: the inductive type function 'dv -> 'sv
its GADT solution. 
its type-level function solution: incope1.hs 
\begin{code}
data P t where
    VI :: Int -> P Int
    VB :: Bool -> P Bool
    VF :: (P a -> P b) -> P (a->b)
    E  :: C t -> P t

abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x) = x

instance Symantics P where
    int x  = VI x
    bool b = VB b

    -- lam :: (repr a -> repr b) -> repr (a->b)
    lam = VF
    -- app :: repr (a->b) -> repr a -> repr b
    app ef ea = case ef of
                        VF f -> f ea
                        E  f -> E (app f (abstr ea))
    add e1 e2 = case (e1,e2) of
                 (VI n1,VI n2) -> VI nr where nr = n1 + n2
                 _ -> E (add (abstr e1) (abstr e2))
\end{code}

GADT: |E|, Purely dynamic; the other alternatives: |static| code.

We need GADT -- or, to be more precise,
representation that varies with the type. We need a value of the type
|P Int| to have a different representation than the value of the type 
|P (Int->Int)|. GADT give you that (although we don't use the GADT in
their intended meaning for that).

Drawbacks of the GADT solution. GADTs, although work, are still not
quite satisfactory: let us look at the following lines in incope.hs
\begin{code}
    app ef ea = case ef of
                        VF f -> f ea
                        E  f -> E (app f (abstr ea))
\end{code}
The datatype has four constructors, |VI|, |VB|, |VF| and |E|. How come we used
only two of them in the above expression. It is obvious, one may say,
that the constructors |VI| and |VB| just can't occur, because of the
typing consideration. This is obvious to us -- but it is not obvious
to the compiler. The case matching above does look \emph{partial}. That
seems like a tenuous point -- but it is precisely the main point of
the whole tag elimination approach. In a typed tagged interpreter
there are lots of pattern-matching that looks partial but in reality
total. The ultimate goal is to make this totality syntactically
apparent.

How can we do without GADT
Now show the Symantics signature in full, with the 'sv types

In the ideal world, the MetaOCaml signature (or Haskell type class)
Symantics would start with
\begin{code}
  kind tp = int | bool | (->) tp tp
  type ('c, 'dv:tp) repr : tp
  val int : int -> ('c, int) repr
  val bool : bool -> ('c, bool) repr
  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> 
            ('c, 'da -> 'db) repr
  val add : ('c, int) repr -> ('c, int) repr -> 
            ('c, int) repr
\end{code}

Here "repr" is a tp-to-tp function, defined inductively in the case of
partial evaluation:

\begin{code}
  ('c, int   ) repr_pe = VI of int
                       | E of ('c, int) code
  ('c, bool  ) repr_pe = VB of bool
                       | E of ('c, bool) code
  ('c, 'a->'b) repr_pe = VF of ('c,'a) repr -> ('c,'b) repr
                       | E of ('c, 'a->'b) code
\end{code}

Unfortunately, we can't define type-level functions inductively.
Instead we bake |repr_pe| above into the signature Symantics, so repr
takes three type arguments |('c,'sv,'dv)|, where the extra argument 
|'sv| is equal to |('c,'dv) repr_pe.|
It seems Ken has just showed how to define type-level functions
inductively... This shows also how a limited form of GADTs (the ones
used for the sake of type-dependent representations) can be
implemented with just functions.

So we have to augment the full signature of Symantics (from that of
sec.~\ref{encoding}) to
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> 
             ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
           'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
             'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : 
    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
         'da->'db) repr 
    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end
\end{code}

\subsection{MetaOCaml version}

%%  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x

%% We add one item to the language, |dyn_id|, as it
%% is needed in the next section, where we will explain its purpose.

Here is the full code for the MetaOCaml version.
\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = 
        {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  (* generic build - takes a repr constructor, an 
     interpreter function and a compiler function (all 
     binary) and builds a PE version *)
  let build cast f1 f2 = fun e1 -> fun e2 -> 
     match (e1,e2) with
         ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))
  (* same as 'build' but takes care of the neutral element 
     (e) simplification allowed via a monoid structure which is 
     implicitly present *)
  let buildsimp cast e f1 f2 = fun e1 e2 -> 
     match (e1,e2) with
        | ({st = Some e'}, _) when e = e' -> e2
        | (_, {st = Some e'}) when e = e' -> e1
        | ({st=Some n1}, {st=Some n2})-> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))

  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
  let leq e1 e2 = 
     match (e1,e2) with
         ({st=Some n1},{st=Some n2})-> bool (R.leq n1 n2)
        | _ -> pdyn (C.leq (abstr e1) (abstr e2))
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (C.dyn_id (abstr ef)) (abstr ea))

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x))) in
    let fdynn e = pdyn (C.app fdyn e.dy) in
    {st=Some (function {st = Some _} as e -> 
               let rec self n = 
                 (match n.st with Some _-> app (f (lam self)) n 
                                 | _ -> fdynn n) in 
               app (f (lam self)) e
             | e  -> fdynn e);
    dy = fdyn }
end;;
\end{code}

\oleg{Compare the output of some of the ctest with the corresponding
ptest. Use power?}

@@

\jacques{Comment on the various nice generic things in the implementation,
which are a little silly in this small version, but would really pay off
in something bigger.  See the abstract interpretation hard at work!
See the monoid-driven simplifications!}
\jacques{Also need to note that the type of the above is not quite
the original Symantics (although the code has the right operational
semantics).  We need to make some adjustments, as per below.}

Also, unlike the GADT
solution, there is no longer any seemingly partial
pattern-matching. The pattern-match only occurs in PE (to check if the
code is static or dynamic), it is only a phase rather than a type tag
match, and the pattern match is \emph{patently} exhaustive. That is,
totality is assured and it is assured \emph{syntactically}, in the way that
is apparent to the compiler.

Say that when doing PE on fix, we made a deliberate decision to `go
all the way' -- so long as the computation can proceed at the PE stage
(`static value'), we do that rather than residualizing. We have also
an alternative where fix is unrolled only once and them we
residualize. We have used the first approach, even it (unlike the
latter) may cause non-termination at the PE stage -- but \emph{only} if the
computation we are partially evaluating is non-terminating and enough
static information is available to trigger this non-termination.
Say that our treating of fix may lead to the code bloat in the
residual program. That is an issue investigated elsewhere (cite our
PEPM2006 paper).

@@

Our approach showed better the meaning of GADT and when they can be
avoided.

>>>>

\section{Base language and its tagless interpreters}\label{language}

Our DSL language to be embedded is simply-typed lambda-calculus
with fixpoint, integers, booleans and comparison, whose BNF grammar is

@@

This time we use higher-order abstract syntax (cite: but not
Elliott. Cite Dale Miller!) rather than de Bruijn indices.  The
language is close to the language of Xi03, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Gibonacci, factorial, and power.

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We have embedded the above language in (Meta)OCaml and Haskell.  In Haskell,
the constructors of DSL terms are functions defined as methods in a type class
Symantics. The name implies that the class interface gives the syntax for
the source language and its instances give the semantics.

@@

We also embedded our DSL in (Meta)OCaml, replacing type classes with
functors, via a module type.  For this section, the following simple
signature will be sufficient:

@@

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
     if_ (leq n (int 0)) (fun () -> int 1)
         (fun () -> mul x (app self 
                               (add n (int (-1))))))))
 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
The dummy unit argument in |testpowfix| is an artefact of MetaOCaml
(which we will be using to compile our tests in Sections below),
again related to monorphism. To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means, the type
must be polymorhic. The value restriction then dictates 
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

\subsection{Two tagless interpreters}
The first instantiation of the Symantics structure is an interpreter:
evaluating the object (DSL) term to its value in the meta-language. 
The following module essentially interprets (gives one-to-one
correspondence to) each operation of the object language as the
corresponding operation in the meta-language.
\begin{code}
module R  = struct
  type ('c,'dv) repr = 'dv (* absolutely no wrappers *)
  let int  (x:int)  = x
  let bool (b:bool) = b
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then (et ()) else (ee ())

@@

We have also implemented the ``compiler'' in Haskell. Since Haskell
has no (convenient, and typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with the
constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
meaning. Unlike our staging language (which is based on higher-order
abstract syntax), our bytecode, to be realistic, uses integer-named
variables. We then define 
\begin{code}
  newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for the creation of fresh variable
names. We define the compiler by making |C| to be the instance of the
class |Symantics|. The implementation is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying complete code gives more details.

@@

\subsection{Non-functorialness of functions problem}
\label{S:PE-problem}
We want to combine the interpreter |R| and the compiler |C| to produce
an (online) partial evaluator. \oleg{discuss if our PE online or
  offline: here or below. In the latter case refer to the discussion
  from here.} The partial evaluator (PE) is also an interpreter: it
evaluates a term either to immediate (or, `static' or present-stage)
value or a `dynamic', future-stage value, or `code' (of the sort
produced by the compiler C). The interpreter attempts to perform all
possible computations (i.e., computations whose operarands are
present-stage values) as the present-stage. Our first attempt is
straightforward: we introduce
\begin{code}
data P1 t = S1 (R t) (C t) | E1 (C t)
\end{code}
as a `direct sum' of the interpreter |R| and the compiler |C|. The
datatype |P1| has two \emph{phase} (not type!) tags |S1| 
(for static) and |E1| for dynamic. The static value has both the
present-stage and future-stage alternatives. The dynamic value does
not have the present-stage component. We define
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (S1 _ dyn) = dyn
abstr1 (E1 dyn)   = dyn
\end{code}
to extract the future-stage component. We then attempt to define the
interpreter |P1| compositionally -- and it works for the first-order
fragment of our DSL language.

\begin{code}
instance Symantics P1 where
    int  x  = S1 (int x) (int x)
    bool b  = S1 (bool b) (bool b)
    add (S1 n1 _) (S1 n2 _) = int (unR (add n1 n2))
    add e1 e2 = E1 (add (abstr1 e1) (abstr1 e2))
    -- mul and leq are analogous and elided
    if_ (S1 s _) et ef = if unR s then et else ef
    if_ eb et ef = 
       E1 (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code}
Integer and boolean literals are obviously immediate, present-stage
values. The result of adding two values is static only if both values
are. If so, we use the evaluator |R| to add the values; the result is
an essentially integer literal, which can be promoted to the future
stage. If at least one of the operands to |add| is dynamic, we promote
both operands to the future-stage and use the compiler |C| to
`compile' the code for adding the operand expressions.  Because the
static value includes as a component its future-stage variant, and
because we build both components together, we are able to get by
without the polymorphic `lift' operation (that is, lift the
computation of any type from the present-stage to the future
stage). We can do such a lifting only on type-by-type basis (for
booleans and for integers). The problem with the polymorphic lift is
that it inhibits true compilation, to assembler or other similar first
order target language with no polymorphic (run-time reified) types.
The representation of integers in the target language most likely
differs from the representation of booleans or functions,
etc. Therefore, without the intentional type analysis, the polymorphic
lift is not compilable.

When we come to the higher-order fragment of our language, however, we
stumble. Let us consider |lam f|. Here, |f| is a function from a
|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
static value or a dynamic value -- but which? Unlike |add| or |if_|, 
|lam| cannot really examine its operand. The arguments of |add| are 
of type |P1 int|, which we can examine by pattern-matching and so 
determine if they are static or dynamic. The argument of |lam| is a 
higher-order function, which is not introspectible. One may only 
apply |f| to a particular static or dynamic value, and receive another 
value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
applying the same function to different static values may give 
either static or dynamic results (in the example above, applying 
the function to (int 0) gives the static value in return). One
solution is to be conservative and always use the |C| compiler to
compile |f| to the future-stage expression, which we can tag with
E1. But that defeats the purpose of partial evaluation as we may miss
many opportunities for static computations and simplifications, such
as multiplication by zero in the example above. Another solution is to
delay the binding-type-analysis (BTA) until the function |f| is
applied to a particular argument (that is, until the result of |lam f|
appears as the argument of |app|). That means, we have to find a way
to incorporate |f| as it is into the P1 data structure. We would like
to write that when |t| is a particular type |a->b|, that data type
should be 
\begin{code}
data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
               | E1 (C (a->b))
\end{code}
That is, we need a non-parameteric data type, something like a
typecase (ref? Gibbons?).

The problem we encountered here is quite akin to typed CPS
transform. When transforming abstractions, we come across
non-factorialness. The connection with CPS is not an accident, as we
shall see below (in Section XXX).

\subsection{GADT -- not quite tagless PE}
\label{S:PE-GADT}

Two common ways of providing typecase in Haskell is by using either
generalized algebraic data types (GADT) or type-class functions
(implemented via typeclasses with functional dependencies). These
methods are equivalent; here we show the GADT; please see |incope1.ml|
in the accompanying source code for the latter.

We introduce the following GADT
\begin{code}
data P t where
    VI :: Int  -> P Int
    VB :: Bool -> P Bool
    VF :: (P a -> P b) -> P (a->b)
    E  :: C t -> P t
\end{code}
with four data constructors. The constructor |VF| stands out as its
operand is of the type |P a -> P b| rather than |a -> b| as one may
have expected if |P t| were parametric in |t|. As before, we define a
function to extract the future-stage computation from the |P t| value:
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}

We use the method |lam| of the |C| intrepreter to compile the argument
of |lam| into the code value. We may now make |P| the instance of
Symantics and thus implement the partial evaluator (the
implementations of |mul|, |leq|, |if_| and |fix| are elided).
\begin{code}
instance Symantics P where
    int x  = VI x
    bool b = VB b

    add e1 e2 = case (e1,e2) of
                 (VI n1,VI n2) -> VI (n1 + n2)
                 _ -> E (add (abstr e1) (abstr e2))
    lam = VF
    app ef ea = case ef of
                        VF f -> f ea
                        E  f -> E (app f (abstr ea))
\end{code}

The implementations of |int|, |bool|, and |add| are similar to the
above. The interpretation of |lam f| merely incorporates the
HOAS function |f| into the result. At any point we can convert it to a
code value, if we must. But delaying such a conversion, we get the
opportunity to apply |f| to a concrete argument. When we evaluating
|app ef ea|, we check to see of the first argument is so-delayed
lambda-expression. If it is, we apply the incorporated |f| to the
concrete argument |ea|, giving us a chance to perform the static
computations (we see the example in the next subsection). If |ef| is a
dynamic value, we residualize.

Although GADT solve our problem, the solution is not quite
satisfactory. First of all, it cannot be ported to MetaOCaml as GADTs
are not supported in OCaml. Second, the tagging problem reappeared, in
the implementation of |app| above. The datatype |P t| has four data
constructors; the pattern-matching in |app| however used only two of
them, |VF| and |E|. It is obvious, one may say, that the constructors
|VI| and |VB| just cannot occur, because the corresponding values
cannot have the arrow type. What is obvious to us is not obvious to
the compiler. The case matching above does look \emph{partial}. The
point may loop tenuous -- and yet it is precisely the main point of
tag elimination. In a typed tagged interpreter there are many
pattern-matching forms that looks partial but are total in
reality. The goal is to make this totality \emph{syntactically}
apparent.

\subsection{The `final' solution}
\label{S:PE-solution}
Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = (repr_pr a) -> (repr_pe b)
\end{code}
Although we can use typeclasses to define that type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with the existing
methods of |Symantics| and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying |repr_pe| function above to the
type |'dv|. The signature |Symantics| now reads
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> 
             ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
           'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
             'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : 
    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
         'da->'db) repr 
    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end;;
\end{code}
The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would have write, in the case of the |C| interpreter,
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of that
type argument |'sv|. Here's the corresponding instance of the
Symantics:

\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = 
        {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  let buildsimp cast e f1 f2 = fun e1 e2 -> 
     match (e1,e2) with
        | ({st = Some e'}, _) when e = e' -> e2
        | (_, {st = Some e'}) when e = e' -> e1
        | ({st=Some n1}, {st=Some n2})-> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))

  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
  let leq e1 e2 = 
     match (e1,e2) with
         ({st=Some n1},{st=Some n2})-> bool (R.leq n1 n2)
        | _ -> pdyn (C.leq (abstr e1) (abstr e2))
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (abstr ef) (abstr ea))

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x))) in
    let fdynn e = pdyn (C.app fdyn e.dy) in
    {st=Some (function {st = Some _} as e -> 
               let rec self n = 
                (match n.st with Some _->app (f (lam self)) n 
                                | _ -> fdynn n) in 
               app (f (lam self)) e
             | e  -> fdynn e);
    dy = fdyn }
end;;
\end{code}

The type |repr| here literally expresses the type equation at the
beginning of this section. The function |abstr|, as in the previous
subsection, extracts the future-stage computation from the result of
PE. The function |pdyn| does the converse. As in
section~\ref{S:PE-problem}, we build future-stage computations as we
build present-stage ones, if any. Therefore, we avoid the need for the
polymorphic `lift' function. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage of the
implicit monoid structure. The function |buildsimp| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function also checks if one argument is static
and is the neutral element of the monoid structure 
\jacques{Comment on the various nice generic things in the implementation,
which are a little silly in this small version, but would really pay off
in something bigger.  See the abstract interpretation hard at work!
See the monoid-driven simplifications!}
\oleg{I'm inclined to ditch buildsimp. We miss an opportunity here:
  when multiplying anything by 0, the result is zero!}

When doing PE on fix, we made a deliberate decision to `go all the
way' -- so long as the computation can proceed at the PE stage
(`static value'), we do that rather than residualizing. In the
accompanying source code, we have also implemented an alternative
where fix is unrolled only once and them we residualize. We prefer the
first approach, even though it (unlike the latter) may cause
non-termination at the PE stage -- but \emph{only} if the computation
we are partially evaluating is non-terminating and enough static
information is available to trigger this non-termination.  Our
treatment of fix may lead to the code bloat in the residual
program. That is an issue investigated elsewhere (cite our PEPM2006
paper).

Given the above Symantics, our running test now evaluates
\begin{code}
  let module E = EX(P) in E.test1 ();;
\end{code}
to
\begin{code}
- : ('a, bool, bool) P.repr = 
   {P.st = Some true; P.dy = .<(true)>.}
\end{code}
It is instructive to compare this result with that of the |C|
evaluator. We observe the beta-redex that was apparent there is now
performed. The result is the static value |true|.
More interesting is the partial evaluation of testpowfix7. The |C|
interpreter evaluates |testpowfix7| to the future-stage code with many
apparent beta-redexes.
\begin{code}
  let module E = EX(C) in E.testpowfix7;;
 - : ('_a, int -> int) C.repr =
.<fun x_1 ->
   (((fun x_2 ->
       let rec self_3 =
        fun n_4 ->
         ((fun x_5 -> if (x_5 <= 0) then 1 
                      else (x_2 * (self_3 (x_5 + (-1)))))
           n_4) in
       self_3) x_1) 7)>.
\end{code}
The result of the |P| interpreter is different:

\begin{code}
  let module E = EX(P) in E.testpowfix7;;
 - : ('_a, ('_a, int, int) P.repr -> ('_a, int, int) P.repr, 
      int -> int) P.repr = 
 {P.st = Some <fun>;
  P.dy = .<fun x_1 -> 
     (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1))))))>.}
\end{code}

which is what one would have expected from the power example.

Unlike the GADT approach in the previous subsection, there is no
longer any seemingly partial pattern-matching. The pattern-match only
occurs in PE (to check if the code is static or dynamic), it is only a
phase rather than a type tag match, and the pattern match is
\emph{patently} exhaustive. That is, totality is assured and it is
assured \emph{syntactically}, in the way that is apparent to the
compiler.

@@

@@

Our approach showed better the meaning of GADT and when they can be
avoided.
This shows also how a limited form of GADTs (the ones
used for the sake of type-dependent representations) can be
implemented with just functions.
